---
title: "WineQuality"
author: "Suchitra"
date: "2/18/2017"
output:
  html_document: default
  pdf_document: default
---
#We will load the dplyr, ggplot and lattice libraries
#dply is used for data manipulation.
#ggplot is used for creating graphics and plots.
#Lattice package is used for high- level data visualization.

```{r setup, include=FALSE}
library(lattice)
library(dplyr)
library(ggplot2)
```

--->Variable definition:

Fixed acidity: Acid involved with wine that are non-volatile.
Volatile acidity: Acetic acid in wine that at high levels leads to a taste of vinegar.
Citric acid: Present in small quantities, add freshness and flavor to wine.
Residual sugar: Sugar thats left after the fermentation stops.
chlorides:The amount of salts in wine.
Free sulphur dioxide:Prevents microbial growth and oxidation of wine.
Total: Amount of free + bound forms of SO2.
Density: Depends on the percent alcohol and sugar content.
PH : Acidic or basic.
Sulphates: Wine additive which contributes to sulphur dioxide.
Alcohol: Percent alcohol content in wine.
Quality: Output variable.

--->Questions:
1. How is human tasting of wine related to the chemical properties.


# Getting familiar with the data.

```{r cars}
wine_data <- read.csv("wineQualityReds.csv", header = TRUE, sep=",")
summary(wine_data)
names(wine_data)
str(wine_data)  
```

---->Variable Identification:
All the variables are categorical or continuous variables. The chemical properties(continuous variables) are predictor variables and quality.(ordinal variable) is the outcome/target variable.

## Univariate Data Analysis.
#We analyze the variables individually to understand their behavior

```{r pressure, echo=FALSE}
#Fixed acidity
ggplot(aes(x= wine_data$fixed.acidity), data = wine_data)+
  geom_histogram()+
  scale_x_continuous(limits = c(6,12))

#Volatile acidity
ggplot(aes(x= wine_data$volatile.acidity), data = wine_data)+
  geom_histogram()+
  scale_x_continuous(limits=c(0.2,0.8))

#Citric acidity
ggplot(aes(x=wine_data$citric.acid), data= wine_data) +
  geom_histogram()+
  scale_x_continuous(limits = c(0,0.6))

#Residual Sugar
ggplot(aes(x=wine_data$residual.sugar), data= wine_data) +
  geom_histogram()+
  scale_x_continuous(limits = c(1,4))

#Chlorides
ggplot(aes(x=wine_data$chlorides), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits = c(0.025,0.15))

#Free suplhu dioxide
ggplot(aes(x=wine_data$free.sulfur.dioxide), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits = c(0,40))

#Total Sulphur dioxide
ggplot(aes(x=wine_data$total.sulfur.dioxide), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits=c(0,100))

#Density
ggplot(aes(x=wine_data$density), data=wine_data)+
  geom_histogram()+
  scale_x_continuous()

#PH
ggplot(aes(x=wine_data$pH), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits = c(3,3.75))

#Sulphates
ggplot(aes(x=wine_data$sulphates), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits=c(0.25,1))

#Alcohol
ggplot(aes(x=wine_data$alcohol), data= wine_data)+
  geom_histogram()+
  scale_x_continuous(limits = c(9,13))

#Quality
ggplot(aes(x=wine_data$quality), data= wine_data)+
  geom_histogram()+
  scale_x_continuous()

```


# Summary Statistics

```{r}
summary(wine_data$fixed.acidity)
summary(wine_data$volatile.acidity)
summary(wine_data$citric.acid)
summary(wine_data$residual.sugar)
summary(wine_data$chlorides)
summary(wine_data$free.sulfur.dioxide)
summary(wine_data$total.sulfur.dioxide)
summary(wine_data$density)
summary(wine_data$pH)
summary(wine_data$sulphates)
summary(wine_data$alcohol)
summary(wine_data$quality)
```


#Correlation

```{r}
cor(x=wine_data[,2:13], y=wine_data[,2:13])

pairs(~wine_data$fixed.acidity+wine_data$volatile.acidity+wine_data$citric.acid+wine_data$residual.sugar+wine_data$chlorides+wine_data$free.sulfur.dioxide+ wine_data$total.sulfur.dioxide+wine_data$density+wine_data$pH+wine_data$sulphates+wine_data$alcohol+wine_data$quality, data=wine_data, main="Scatter plots")
```
Quality of wine correlates strongly with alcohol, volatile acididty, suplhates, citric acid, as compared to the other variables. So we can consider these four variables to understand their association with wine quality scores.


# Scatterplot matrix
# Lets consider the highest correlated chemical properties.

#Alcohol
```{r}
ggplot(aes(x=wine_data$quality, y= wine_data$alcohol), data= wine_data) +
  geom_jitter(alpha=1/10)+
  scale_x_continuous(breaks = seq(0,8,1))+
  geom_smooth(method='lm', aes(group=1))+
  geom_hline(yintercept = mean(wine_data$alcohol), linetype="longdash", color='red')+
  geom_vline(xintercept = mean(wine_data$quality), linetype="longdash", color='red')
```

Relationship between alcohol and wine scores:
From the scatterplot, we can see that the wine scores have a positive relationship with the percent alcohol content in wine. Wine scores in the higher range of 5 to 8, tend to have a higher alcohol content between 10 to 12.

#Volatile acidity
```{r}
tapply(wine_data$volatile.acidity, wine_data$quality, mean)
ggplot(aes(x= wine_data$quality, y=wine_data$volatile.acidity), data= wine_data)+
  geom_jitter(alpha=1/10)+
  geom_smooth(method='lm', aes(group=1))+
  geom_hline(yintercept = mean(wine_data$volatile.acidity), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$quality), color='red', linetype='longdash')
```
Relationship between volatile acidity and wine scores.
From the plot, we can say that wines have a lower score for higher values of volatile acidity. This indicates that they have a negative association. This might be intuitive as volatile acidity in wine increases the acetic acid content and leads to a more prominent vinegar taste. Presence of this taste is a not a good indication of a high quality wine.


#Sulphates

```{r}
tapply(wine_data$sulphates, wine_data$quality, mean)
ggplot(aes(x= wine_data$quality, y=wine_data$sulphates), data= wine_data)+
  geom_jitter(alpha=1/10)+
  geom_smooth(method='lm', aes(group=1))+
  geom_hline(yintercept = mean(wine_data$sulphates), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$quality), color='red', linetype='longdash')
```
Relationship between suplhates and wine qulaity scores:
There seems to a be a positive asscociation between wine quality scores and sulphates content in wine. As suplhates prevents microbial growth and oxidation of wine, it might be a good indication of a high quality wine. The wine quality score increases drastically with just a small increase in sulphates.


# Citric acid
```{r}
tapply(wine_data$citric.acid, wine_data$quality, mean)
ggplot(aes(x= wine_data$quality, y=wine_data$citric.acid), data= wine_data)+
  geom_jitter(alpha=1/10)+
  geom_smooth(method='lm', aes(group=1))+
  geom_hline(yintercept = mean(wine_data$citric.acid), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$quality), color='red', linetype='longdash')
```
Realtionship between wine qulaity scores and citric acid content:
There seems to be a positive relationship between the citric acid content and the wine quality scores. As citric acid brings freshness to the taste of wine, it might be indicative of a high qulaity wine.



# Multivarate data analysis:
# Volatile acidity vs alcohol
```{r}
ggplot(aes(wine_data$alcohol, wine_data$volatile.acidity), data=wine_data)+
  geom_jitter(aplha=1/20)+
  geom_point(aes(color=quality)) +
  geom_hline(yintercept = mean(wine_data$volatile.acidity), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$alcohol), color='red', linetype='longdash')
```
From the plot, we can see that wines having high alcohol content and low volatile acidity have higher quality scores. But there is noe specific values of alcohol content and volatile acidity for lower quality scores.

#Citric acid vs alcohol
```{r}
ggplot(aes(wine_data$alcohol, wine_data$citric.acid), data=wine_data)+
  geom_point(aes(color=quality)) +
  geom_hline(yintercept = mean(wine_data$citric.acid), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$alcohol), color='red', linetype='longdash')
  

```
From the plot, we can see that wines have higher quality scores, when their alcohol and citric acid content are in the higher ranges. Again there is noe specific pattern for low quality scores of wine.

#Citric acid vs volatile acidity
```{r}
ggplot(aes(wine_data$volatile.acidity, wine_data$citric.acid), data=wine_data)+
  geom_point(aes(color=quality)) +
  geom_hline(yintercept = mean(wine_data$citric.acid), color='red', linetype='longdash')+
  geom_vline(xintercept = mean(wine_data$volatile.acidity), color='red', linetype='longdash')
```
From the plot we can see that wines having a higher level citric content and lower level volatile acid have higher quality scores.



#Building a linear regression model.
```{r}
quality_mdl <- lm(wine_data$quality ~ wine_data$alcohol +wine_data$volatile.acidity + wine_data$sulphates +wine_data$citric.acid, data=wine_data)

summary(quality_mdl)
```
Intercept is 1.874
slope is 0.36084
Thus the formula is: quality= 0.36*alcohol + 1.874
Both the Pr values for the intercept and the slope is less than 0.05.The stars at the right of these value indicate the level of significance.
p value implies that the observtion is statistically significant.


#Testing the Linear regression assumptions with the help of the residual diagnostics
#Residual analysis.
```{r}

xyplot(resid(quality_mdl) ~ fitted(quality_mdl),
  xlab = "Fitted Values",
  ylab = "Residuals",
  main = "Residual Diagnostic Plot",
  panel = function(x, y, ...)
  {
    panel.grid(h = -1, v = -1)
    panel.abline(h = 0)
    panel.xyplot(x, y, ...)
  }
)
```
We plot the predicted values which we get using the quality_mdl on the x axis and the residual values on the y axis. Ecah point in this plot is a wine quality score, with the predicted values on the x axis and the accuracy on the y axis.
The residual diagnostic plot shows a peculian pattern, and they are not evenly distributed. Theere re no clusters at the middle of the plot or around the lower values of y. Thus this signifies that  there is significant room for improvement in the model that we built.

#Prediction
```{r}
#splitting data 75:25

#Computing sample_size of the train dataset
sample_size <- floor(0.7*nrow(wine_data))
#Load the train and test data
set.seed(100)
train_indices <- sample(seq_len(nrow(wine_data)), size = sample_size)


#Load the train and test dataset
train_data <- wine_data[train_indices,]
test_data <- wine_data[-train_indices,]

#Build a prediction model
linear1 <- lm(wine_data$quality ~ wine_data$alcohol+wine_data$sulphates+wine_data$citric.acid+wine_data$volatile.acidity, train_data)
summary(linear1)

#Testing the prediction model
predicted=predict(linear1,test_data)
head(predicted)
head(test_data$quality)

#Calculating multiple R squared for test data
SSE <- sum((test_data$quality-predicted)^2)
SST <-sum((test_data$quality-mean(test_data$quality))^2)
1-(SSE/SST)
```
Here the multiple R squared value denotes the variance explained by the model. Since R^2 is less than zero, we can say that this model is a bad fit to the data.

We now use a random forest regression model on the wine quality data, to see if the model is a good fit and make preictions for the quality of the produce.

#Random forest
```{r}
library(randomForest)
wine_rf <- randomForest(as.factor(quality)~alcohol+sulphates+citric.acid+volatile.acidity, data=train_data, importance=TRUE, ntree=1000,mtry=2)
varImpPlot(wine_rf)


wine_rf

wine_predict <- predict(wine_rf,test_data)
table(wine_predict, test_data$quality)
wine_acc <- (170+118+30)/nrow(test_data)
wine_acc
```
We have built a random forest model using the 4 important variables. The variable importance plot shows the effect of variable removal on the accuracy of the model. Thus the alcohol content in wine predicts the quality score the best, followed by volatile acidity, sulphates and then citric acid.

The random forest model fit on the wine quality data, gives an accuracy of 0.66, which high. 